---
title: "Types of 1D Spline"
author: "Lindesay Scott-Hayward"
date: "2023-01-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document shows how to fit both one and two dimensional splines.

There are three different types of 1D spline:

- B-spline
- Cyclic Cubic Spline
- Natural Cubic Spline


and two different 2D radial basis functions

- exponential
- gaussian

# One dimensional Smoothing

## Starting point

- Load some data
- specify the variable of interest. In this case we pick month of the year. 
- Fit an initial model.  For simplicity we fit an intercept only model.

```{r}
# load data
data(ns.data.re)
```

```{r}
ns.data.re$response<- ns.data.re$birds
```

```{r}
varlist=c('MonthOfYear')
```

```{r}
initialModel<- glm(response ~ 1, family='quasipoisson',data=ns.data.re)
```



## Default: B-spline

- This section shows the fitting of a quadratic B-spline.  If linear or cubic B-splines are preferred, change `degree` to 1 or 3 respectively. 
- The default setting is for B-splines. If the `splines` parameter in the `salsa1dlist` is not specified then B-splines will be fitted.
- Here I have shown how to add the `splines` parameter to the `salsa1dlist` object. 

Note:  I have chosen the QBIC for chosing the number and location of knots as it is relatively quick and useful for testing. (Q because we have a Quasi-Poisson model).

```{r}
# set some input information for SALSA
salsa1dlist<-list(fitnessMeasure = 'QBIC', 
                  minKnots_1d = c(1), 
                  maxKnots_1d = c(3), 
                  startKnots_1d = c(1), 
                  degree = c(2),
                  gaps = c(0),
                  splines = c("bs"))
```


```{r}
# run SALSA
salsa1dOutput<-runSALSA1D(initialModel, 
                          salsa1dlist, 
                          varlist = varlist, 
                          datain = ns.data.re,
                          suppress.printout = TRUE)

```


```{r}
runPartialPlots(salsa1dOutput$bestModel, data=ns.data.re, varlist.in = varlist, showKnots = TRUE)
```



## Cyclic spline

To fit a cyclic spline, the splines parameter is set to "cc".  To change the order of a cyclic spline, you can use the `degree` parameter (order = degree +1).

Note that even though the minimum knots is set to 1, for a cyclic spline the minimum allowed is 3.  This is changed automatically. 


```{r}
#set some input info for SALSA
salsa1dlist<-list(fitnessMeasure = 'QBIC', 
                  minKnots_1d = c(1), 
                  maxKnots_1d = c(3), 
                  startKnots_1d = c(1), 
                  degree = c(2),
                  gaps = c(0),
                  splines = c("cc"))
```


```{r}
# run SALSA
salsa1dOutput.cc<-runSALSA1D(initialModel, 
                          salsa1dlist, 
                          varlist = varlist, 
                          datain = ns.data.re,
                          suppress.printout = TRUE)

```


```{r}
runPartialPlots(salsa1dOutput.cc$bestModel, data=ns.data.re, varlist.in = varlist, showKnots = TRUE)
```



## Natural Cubic spline

To fit a natural spline, the splines parameter is set to "ns".

Note that `degree` is not necessary for this spline type but it must still be specified in the `salsa1dlist` object.  You can use NA for this. 

```{r}
#set some input info for SALSA
salsa1dlist<-list(fitnessMeasure = 'QBIC', 
                  minKnots_1d = c(1), 
                  maxKnots_1d = c(3), 
                  startKnots_1d = c(1), 
                  degree = c(NA),
                  gaps = c(0),
                  splines = c("ns"))
```


```{r}
# run SALSA
salsa1dOutput.ns<-runSALSA1D(initialModel, 
                          salsa1dlist, 
                          varlist = varlist, 
                          datain = ns.data.re,
                          suppress.printout = TRUE)

```


```{r}
runPartialPlots(salsa1dOutput.ns$bestModel, data=ns.data.re, varlist.in = varlist, showKnots = TRUE)
```

## Multiple different splines

To fit more than one type of spline to different variables, this is specified in the `splines` parameter in the same way you would specify the starting knots for example. 

```{r}
varlist=c('MonthOfYear', "x.pos")
```

```{r}
#set some input info for SALSA
salsa1dlist<-list(fitnessMeasure = 'QBIC', 
                  minKnots_1d = c(1,1), 
                  maxKnots_1d = c(3,3), 
                  startKnots_1d = c(1,1), 
                  degree = c(2, 2),
                  gaps = c(0, 0),
                  splines = c("cc", "bs"))
```


```{r}
# run SALSA
salsa1dOutput.multi<-runSALSA1D(initialModel, 
                          salsa1dlist, 
                          varlist = varlist, 
                          datain = ns.data.re,
                          suppress.printout = TRUE)

```


```{r}
runPartialPlots(salsa1dOutput.multi$bestModel, data=ns.data.re, varlist.in = varlist, showKnots = TRUE)
```

# Two dimensional Smoothing

## Starting point

This is the same starting point for one dimensional splines if it is only a two dimensional smooth you want. 

- Load some data
- specify the variable of interest. In this case we pick month of the year. 
- Fit an initial model.  For simplicity we fit an intercept only model.

```{r}
# load data
require(dplyr)
baselinedata <- filter(nysted.analysisdata, impact == 1, season == 1)
```

```{r}
initialModel<- glm(response ~ 1, family='quasipoisson',data=baselinedata)
```


```{r}
kg <- getKnotgrid(baselinedata[, c("x.pos", "y.pos")], numKnots = 300, plot = TRUE)
```

```{r}
# make distance matrices for datatoknots and knottoknots
distMats<-makeDists(baselinedata[, c("x.pos", "y.pos")], kg)
```

```{r}
# make prediction distance matrix. 
preddata <- filter(nysted.predictdata, impact == 0, season == 1)
p2k <-makeDists(preddata[, c("x.pos", "y.pos")], kg, knotmat = FALSE)$dataDist
```



## Default: Gaussian Radial Basis

This uses a gausian radial basis implemented using `LRF.g()`.  It is the default and does not need to be specified. 

```{r}
# make parameter set for running salsa2D
salsa2dlist<-list(fitnessMeasure = 'QBIC', 
                  knotgrid = na.omit(kg),
                  startKnots=10, 
                  minKnots=2, 
                  maxKnots=20, 
                  gap=0)
```

```{r}
salsa2dOutput<-runSALSA2D(initialModel,
                          salsa2dlist, 
                          d2k=distMats$dataDist,
                          k2k=distMats$knotDist,
                          basis = "gaussian", ##
                          suppress.printout = TRUE)

```


```{r}
library(ggplot2)
preddata$preds.g <- predict(object = salsa2dOutput$bestModel, newdata = preddata, g2k = p2k)

ggplot() +
  geom_tile(data=preddata, aes(x.pos, y.pos, fill=preds.g, height=sqrt(area), width=sqrt(area))) + 
  xlab("Easting (km)") + ylab("Northing (km)") + coord_equal() +
  theme_bw() + ggtitle("Gaussian Basis") +
  scale_fill_distiller(palette = "Spectral",name="Animal Counts")
```

## Exponential Basis

This uses an exponential radial basis implemented using `LRF.e()`.  It is more peaked at the knot than the gaussian and for this reason I have used a larger number of start knots. 

```{r}
# make parameter set for running salsa2D
salsa2dlist<-list(fitnessMeasure = 'QBIC', 
                  knotgrid = na.omit(kg),
                  startKnots=10, ##
                  minKnots=2, 
                  maxKnots=20, 
                  gap=0)
```

```{r}
salsa2dOutput.exp<-runSALSA2D(initialModel,
                          salsa2dlist, 
                          d2k=distMats$dataDist,
                          k2k=distMats$knotDist,
                          basis = "exponential", ##
                          suppress.printout = TRUE)

```

```{r}
preddata$preds.e <- predict(object = salsa2dOutput.exp$bestModel, newdata = preddata, g2k = p2k)

ggplot() +
  geom_tile(data=preddata, aes(x.pos, y.pos, fill=preds.e, height=sqrt(area), width=sqrt(area))) + 
  xlab("Easting (km)") + ylab("Northing (km)") + coord_equal() +
  theme_bw() + ggtitle("Exponential Basis") +
  scale_fill_distiller(palette = "Spectral",name="Animal Counts")

```
## Raw data for reference

```{r}
ggplot() +
     geom_tile(data=baselinedata, aes(x.pos, y.pos, fill=response, height=sqrt(area), width=sqrt(area))) + 
     xlab("Easting (km)") + ylab("Northing (km)") + coord_equal() +
     theme_bw() + ggtitle("Raw Data") +
     scale_fill_distiller(palette = "Spectral",name="Animal Counts")
```

